Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-04 paper/galsize_paper.tex
Reorganized subsections

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/baseline_sham_orphans.pdf}
\caption{
{\bf SHAM + orphan clustering predictions.}
Using the abundance matching methods described in \S\ref{subsec:smhm}, we compare the projected clustering of mock vs. SDSS galaxies. Each panel shows the comparison for a volume-limited sample defined by a different $\mstar-$threshold. Black points with error bars show SDSS measurements; solid red (black) curves show the abundance matching prediction including (excluding) the effect of orphan subhalos (see Appendix A). Including orphans mitigates the discrepancy in the clustering predicted by traditional, $\mpeak-$based SHAM based, though mild tension remains for $\mstar\gtrsim10^{10.75}\msun.$
}
\label{fig:baseline_sham_clustering}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

\subsection{Size-Dependent Clustering}
\label{subsec:clustering_results}

@@

\subsection{Central vs. Satellite Sizes}
\label{subsec:censat_sizes}

The clustering measurements shown in the previous section present a puzzle. The modeling assumption $\rhalf\propto\rvir$ predicts size-dependent galaxy clustering that is in quite good agreement with observations, particularly considering the simplicity of the model. Yet, small galaxies cluster more strongly relative to large, which seems at odds with naive expectations based on halo mass, since $\rvir\propto\mvir.$ A straightforward resolution to this puzzle is shown in Figure \ref{fig:censatsizehist}, which compares the $\rhalf$ distributions of central, satellite, and splashback galaxies with the same halo mass $\mhalo\equiv\mpeak\approx10^{12}\msun.$ A ``splashback central"  is defined as a present-day central that used to be a satellite, i.e., its main progenitor halo passed inside the virial radius of a larger halo at some point in its past history. On the other hand, we define a ``true central" as a galaxy that has never been a satellite.

In the $\rvir-$based model, satellite and splashback galaxies are smaller than centrals of the same halo mass due to the physical size of their halo being smaller at earlier times $\zpeak.$ There are two distinct reasons why this feature results in small galaxies being more strongly clustered relative to larger galaxies of the same mass. First and foremost, satellite galaxies statistically occupy higher mass host halos that are more strongly clustered. So in models where satellites are smaller than centrals, at fixed $\mstar$ the ``small" subsample will naturally have a higher satellite fraction, boosting the clustering of small galaxies relative to large of the same stellar mass. Second, at fixed mass, halos of $L_\ast$ galaxies that form earlier are more strongly clustered, a phenomenon commonly known as {\em halo assembly bias}. Splashback halos are typically earlier-forming than true centrals, and so models where splashback halos host smaller-than-average galaxies will naturally predict smaller galaxies being the more strongly clustered. We highlight this point in section \ref{subsec:orphan_stripping} below by testing predictions for $\mstar-$based models that do not possess central vs. satellite size differences.

\subsection{Failure of $\mstar-$based Models}
\label{subsec:orphan_stripping}

@@

The treatment of stellar mass loss in this model naturally results in an enhancement of size differences between satellites and centrals. As shown by the purple curves in Figure \ref{fig:mstarmodelclustering}, these central/satellite differences impact $\rhalf-$dependent clustering in the expected manner: small galaxies cluster more strongly relative to large galaxies of the same stellar mass. Since the satellite fraction increases with decreasing stellar mass, the clustering ratios are more negative for samples with smaller $\mstar$ thresholds. This confirms the notion that central vs. satellite differences drive the $\rhalf-$dependence of clustering, although the magnitude of the effect is evidently is not strong enough to produce clustering predictions that are consistent with SDSS. It is difficult to strip enough mass from satellites so that the $\rhalf-$dependent clustering is in agreement with observations. Taken together with the results for the $\rvir-$based model presented in \S\ref{subsec:clustering_results} and shown in Figure \ref{fig:clustering_ratio_upshot}, this supports the conclusion that the relative difference between central and satellite size is at least partially in place at the time of satellite infall. We discuss the physical implications of this result in \S\ref{subsec:satellite_discussion}.

@@

The contrast between the two bottom panels is stark. The clustering of large vs. small bulge-dominated galaxies has almost no dependence upon $\rhalf.$ On the other hand, for disk-dominated galaxies, small galaxies are much more strongly clustered, and the effect is strong. In fact, comparing the bottom left panel of Figure \ref{fig:colorclustering} to the top right panel of Figure \ref{fig:rvir_only_clustering_absolute}, we see that $\rhalf-$dependent clustering is quite similar between disk-dominated galaxies and $\mstar-$threshold samples. We discuss a halo model interpretation of these trends in \S\ref{subsec:future}.

@@

\section{Discussion}
\label{sec:discussion}

\subsection{Relation to Previous Work}
\label{subsec:previous_work}

@@

As discussed in \S\ref{subsec:future}, we are currently pursuing follow-up work in which we jointly model morphological type together with galaxy size. We consider forward modeling methods to be a requisite for meaningful progress on this issue, not only to properly handle the multi-dimensional nature of the problem, but also to rigorously treat systematic errors that plague inference based on galaxy group catalogs \citep[see][for a thorough discussion]{campbell_etal15}.

Our methodology is closely aligned with \citet{somerville_etal17}, who studied the empirical modeling features that are necessary to recover the tight scatter in the observed $\mean{\rhalf}{\mstar}$ relation. By building models where $\rhalf$ is set by halo spin $\lambda_{\rm halo}$, the authors in \citet{somerville_etal17} found that the level of intrinsic scatter about $\langle\lambda_{\rm halo}\vert\mhalo\rangle$ in dark matter halos is at least as large as the scatter about $\langle\rhalf\vert\mstar\rangle$ seen in observed galaxies. In our approach, the level of scatter is simply a modeling parameter whose fiducial value was motivated by \citet{somerville_etal17}. In ongoing follow-up work discussed in \S\ref{subsec:future}, we will systematically test the large-scale structure implications of the assumption that $\lambda_{\rm halo}\propto\rhalf^{\rm disk}$ using forward-modeling methods analogous to those advocated for in \citet{somerville_etal17} and employed here.

\subsection{Implications for Satellite Evolution}
\label{subsec:satellite_discussion}

\subsection{Future Directions for Empirical Modeling of Galaxy Size}
\label{subsec:future}

\subsubsection{Jointly modeling $\langle\mstar\vert\mhalo\rangle$}

@@

\subsubsection{Jointly modeling morphology}

@@

Encouragingly, when tuning $\mathcal{F}_{\rm orphan}$ so that the overall $\mstar-$threshold clustering is roughly correct (as in Figure \ref{fig:baseline_sham_clustering}), the same level of tuning is required for a successful prediction of $\rhalf-$dependent clustering. This figure gives quantitative motivation for the program described in \S\ref{subsec:future} to jointly constrain the satellite fraction $F_{\rm sat}(\mstar)$ together with models for galaxy size.

>>>>

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/baseline_sham_orphans.pdf}
\caption{
{\bf SHAM + orphan clustering predictions.}
Using the abundance matching methods described in \S\ref{subsec:sham}, we compare the projected clustering of mock vs. SDSS galaxies. Each panel shows the comparison for a volume-limited sample defined by a different $\mstar-$threshold. Black points with error bars show SDSS measurements; solid red (black) curves show the abundance matching prediction including (excluding) the effect of orphan subhalos (see Appendix A). Including orphans mitigates the discrepancy in the clustering predicted by traditional, $\mpeak-$based SHAM based, though mild tension remains for $\mstar\gtrsim10^{10.75}\msun.$
}
\label{fig:baseline_sham_clustering}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

\subsection{Size-Dependent Clustering Measurements}
\label{subsec:clustering_results}

@@

\subsection{Failure of $\mstar-$based Models}
\label{subsec:mstar_stripping}

@@

The treatment of stellar mass loss in this model naturally results in an enhancement of size differences between satellites and centrals. As shown by the purple curves in Figure \ref{fig:mstarmodelclustering}, these central/satellite differences impact $\rhalf-$dependent clustering in the expected manner: small galaxies cluster more strongly relative to large galaxies of the same stellar mass. Since the satellite fraction increases with decreasing stellar mass, the clustering ratios are more negative for samples with smaller $\mstar$ thresholds. This confirms the notion that central vs. satellite differences drive the $\rhalf-$dependence of clustering, although the magnitude of the effect is evidently is not strong enough to produce clustering predictions that are consistent with SDSS. It is difficult to strip enough mass from satellites so that the $\rhalf-$dependent clustering is in agreement with observations. Taken together with the results for the $\rvir-$based model presented in \S\ref{subsec:clustering_results} and shown in Figure \ref{fig:clustering_ratio_upshot}, this supports the conclusion that the relative difference between central and satellite size is at least partially in place at the time of satellite infall. We discuss the physical implications of this result in \S\ref{sec:interpretation}.

@@

The contrast between the two bottom panels is stark. The clustering of large vs. small bulge-dominated galaxies has almost no dependence upon $\rhalf.$ On the other hand, for disk-dominated galaxies, small galaxies are much more strongly clustered, and the effect is strong. In fact, comparing the bottom left panel of Figure \ref{fig:colorclustering} to the top right panel of Figure \ref{fig:rvir_only_clustering_absolute}, we see that $\rhalf-$dependent clustering is quite similar between disk-dominated galaxies and $\mstar-$threshold samples. We discuss a halo model interpretation of these trends in \S\ref{sec:future}.

@@

\section{A Unified Interpretation of $\rhalf-$Dependent Clustering}
\label{sec:interpretation}

The clustering measurements shown in the previous section present (at least) two intriguing puzzles. First, the modeling assumption $\rhalf\propto\rvir$ predicts size-dependent galaxy clustering that is in quite good agreement with observations, particularly considering the simplicity of the model. Yet, small galaxies cluster more strongly relative to large, which seems at odds with naive expectations based on halo mass, since $\rvir\propto\mvir.$ Second, there appears to be a highly non-trivial dependence of $\rhalf-$dependent clustering upon whether or not the galaxy sample has been selected by color or morphology. We propose a simple and unified solution to both puzzles in \S\ref{subsec:censat_sizes}, and discuss the broader implications of our interpretation in \S\ref{subsec:broader_implications}.

\subsection{The Relative Size of Centrals vs. Satellites}
\label{subsec:censat_sizes}

A straightforward resolution to this puzzle is shown in Figure \ref{fig:censatsizehist}, which compares the $\rhalf$ distributions of central, satellite, and splashback galaxies with the same halo mass $\mhalo\equiv\mpeak\approx10^{12}\msun.$ A ``splashback central"  is defined as a present-day central that used to be a satellite, i.e., its main progenitor halo passed inside the virial radius of a larger halo at some point in its past history. On the other hand, we define a ``true central" as a galaxy that has never been a satellite.

In the $\rvir-$based model, satellite and splashback galaxies are smaller than centrals of the same halo mass due to the physical size of their halo being smaller at earlier times $\zpeak.$ There are two distinct reasons why this feature results in small galaxies being more strongly clustered relative to larger galaxies of the same mass. First and foremost, satellite galaxies statistically occupy higher mass host halos that are more strongly clustered. So in models where satellites are smaller than centrals, at fixed $\mstar$ the ``small" subsample will naturally have a higher satellite fraction, boosting the clustering of small galaxies relative to large of the same stellar mass. Second, at fixed mass, halos of $L_\ast$ galaxies that form earlier are more strongly clustered, a phenomenon commonly known as {\em halo assembly bias}. Splashback halos are typically earlier-forming than true centrals, and so models where splashback halos host smaller-than-average galaxies will naturally predict smaller galaxies being the more strongly clustered. We highlight this point in section \ref{subsec:mstar_stripping} below by testing predictions for $\mstar-$based models that do not possess central vs. satellite size differences.

\subsection{Broader Implications}
\label{subsec:broader_implications}

\section{Relation to Previous Work}
\label{subsec:previous_work}

@@

As discussed in \S\ref{sec:future}, we are currently pursuing follow-up work in which we jointly model morphological type together with galaxy size. We consider forward modeling methods to be a requisite for meaningful progress on this issue, not only to properly handle the multi-dimensional nature of the problem, but also to rigorously treat systematic errors that plague inference based on galaxy group catalogs \citep[see][for a thorough discussion]{campbell_etal15}.

Our methodology is closely aligned with \citet{somerville_etal17}, who studied the empirical modeling features that are necessary to recover the tight scatter in the observed $\mean{\rhalf}{\mstar}$ relation. By building models where $\rhalf$ is set by halo spin $\lambda_{\rm halo}$, the authors in \citet{somerville_etal17} found that the level of intrinsic scatter about $\langle\lambda_{\rm halo}\vert\mhalo\rangle$ in dark matter halos is at least as large as the scatter about $\langle\rhalf\vert\mstar\rangle$ seen in observed galaxies. In our approach, the level of scatter is simply a modeling parameter whose fiducial value was motivated by \citet{somerville_etal17}. In ongoing follow-up work discussed in \S\ref{sec:future}, we will systematically test the large-scale structure implications of the assumption that $\lambda_{\rm halo}\propto\rhalf^{\rm disk}$ using forward-modeling methods analogous to those advocated for in \citet{somerville_etal17} and employed here.

\section{Future Directions for Empirical Modeling of Galaxy Size and Morphology}
\label{sec:future}

{\color{blue} APH comment: Some of this discussion can/should be migrated to \S\ref{subsec:broader_implications}. Depending on how that section ends up looking, it may be that this entire Future Directions section can be deleted after cannibalization.}

\subsection{Jointly modeling $\langle\mstar\vert\mhalo\rangle$}
\label{subsec:joint_smhm_modeling}

@@

\subsection{Jointly modeling morphology}
\label{subsec:joint_morphology_modeling}

@@

Encouragingly, when tuning $\mathcal{F}_{\rm orphan}$ so that the overall $\mstar-$threshold clustering is roughly correct (as in Figure \ref{fig:baseline_sham_clustering}), the same level of tuning is required for a successful prediction of $\rhalf-$dependent clustering. This figure gives quantitative motivation for the program described in \S\ref{sec:future} to jointly constrain the satellite fraction $F_{\rm sat}(\mstar)$ together with models for galaxy size.
